---
output:
  html_document: default
  pdf_document: default
---

# Data analysis of the student AI for DS Survey data
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(janitor)
library(ggplot2)
library(rlang)
library(ggmosaic)
library(tidyverse)
library(knitr) 
library(forcats)
library(tidyr)
library(stringr)
```


```{r, warning=FALSE, message=FALSE}


library(readxl)


Student <- read_excel("/Users/line/Downloads/AI4DS-Student-Survey_October 5, 2025_15.53.xlsx")


# Extract short and long names
short <- read_excel("AI4DS-Student-Survey_October 5, 2025_15.53.xlsx", 
                    range = "A1:BN1", col_names = FALSE) |> unlist()

long <- read_excel("AI4DS-Student-Survey_October 5, 2025_15.53.xlsx", 
                   range = "A2:BN2", col_names = FALSE) |> unlist()

# Make sure both match the actual column count
n_cols <- ncol(Student)
short <- short[1:n_cols]
long  <- long[1:n_cols]

# Reload the data, skipping first 2 rows, and apply short names
StudentDF <- read_excel("AI4DS-Student-Survey_October 5, 2025_15.53.xlsx", skip = 2, col_names = short)

# Map short names to long labels
header_labels <- setNames(long, short)

# Filter data
FilteredDF <- StudentDF %>%
  filter(`Duration (in seconds)` >= 120, Progress >= 25)

# Print dimensions + sample size
cat("Remaining sample size:", nrow(FilteredDF), "students\n")
dim(FilteredDF)


```

## set up for the colors
```{r , warning=FALSE, message=FALSE}
library(dplyr); library(tidyr); library(forcats); library(ggplot2); library(scales)
## use this palette for all next figures
palette_redteal <- c("#117A65","#48C9B0","#F1C40F","#E67E22","#8B0000" )



```

## Saving custom theme for future use
```{r}
# --- Define once at the top of your script or in a separate R file ---
theme_pub <- function(base_size = 10) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title      = element_text(face = "bold", size = 11, hjust = 0.5),
      plot.subtitle   = element_text(size = 10, hjust = 0.5, color = "gray25"),
      axis.text.x     = element_text(face = "bold", size = 10),
      axis.text.y     = element_text(face = "bold", size = 10, lineheight = 1.1),
      axis.title.x    = element_text(face = "bold", size = 10),
      axis.title.y    = element_text(face = "bold", size = 10),
      panel.grid.major.x = element_line(color = "gray90"),
      panel.grid.minor = element_blank(),
      legend.title    = element_text(face = "bold"),
      legend.text     = element_text(size = 8),
      legend.position = "bottom",
      legend.key.size = unit(0.35, "cm"),
      legend.margin   = margin(t = 2, b = 5),
      legend.box.margin = margin(5, 5, 10, 5),
      legend.box = "vertical",
      plot.caption    = element_text(size = 9, hjust = 0, color = "gray30"),
      plot.margin     = margin(10, 15, 45, 15)
    )
}

###
load_pub_style <- function() {
  # Color palette
  assign("pal_q", c(
    "Not at All Familiar" = "#B2182B",
    "Not Very Familiar"   = "#E67E22",
    "Neutral"             = "#F1C40F",
    "Somewhat Familiar"   = "#48C9B0",
    "Very Familiar"       = "#117A65"
  ), envir = .GlobalEnv)
  
  # Theme
  theme_set(theme_pub())
}

load_pub_style()

```

## Demographics sex table Q3-6
```{r, warning=FALSE, message=FALSE}
## Demographics table Q3-6
library(dplyr)
library(janitor)
library(ggplot2)
library(scales)
library(knitr)
library(kableExtra)

# ---- 1) Q3 Summary: Student Sex ----
q3_summary <- FilteredDF %>%
  filter(!is.na(Q3)) %>%
  tabyl(Q3) %>%
  adorn_totals("row") %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 3)


# ---- 1) Q3 Summary: Student Sex ----


# total valid responses
n_valid_q3 <- FilteredDF %>% filter(!is.na(Q3)) %>% nrow()

q3_summary <- FilteredDF %>%
  filter(!is.na(Q3)) %>%
  tabyl(Q3) %>%                 # gives Q3 + n
  adorn_totals("row") %>%       # adds Total row
  mutate(
    percent = n / n_valid_q3,   # keep counts in `n`, compute percent separately
    percent = percent(percent, accuracy = 0.1)  # format, e.g., "61.2%"
  )

q3_summary

# Display as a formatted table
kable(q3_summary, caption = "Q3: Student Sex - Counts and Percentages") %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 13)
##

# Optional bar plot
ggplot(FilteredDF %>% filter(!is.na(Q3)), aes(x = Q3)) +
  geom_bar(fill = "#48C9B0", color = "gray30", width = 0.7) +
   coord_flip()+
  geom_text(stat = "count", aes(label = ..count..), vjust = 1, fontface = "bold") +
  labs(
    title = "Student Sex",
    x = NULL, y = "Count"
  ) +
  theme_pub() +
  theme(
    axis.text.x = element_text(face = "bold"),
    plot.margin = margin(10, 20, 25, 15)
  )





```

## major
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(stringr)
library(janitor)
library(ggplot2)
library(kableExtra)
library(forcats)


# ---- 2) Q4 Summary (e.g., Major, Class Level, or Other Demographic) ----
q4_summary <- FilteredDF %>%
  filter(!is.na(Q4)) %>%
  tabyl(Q4) %>%
  adorn_totals("row") %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(digits = 3)

# ---- 1) Clean and consolidate major names ----
cleaned_majors <- FilteredDF %>%
  filter(!is.na(Q4)) %>%
  mutate(
    major_clean = str_to_lower(Q4),  # make lowercase
    major_clean = str_squish(major_clean),  # remove double spaces
    
    # Consolidate common variations
    major_group = case_when(
      str_detect(major_clean, "stat") |
      str_detect(major_clean, "data") |
      str_detect(major_clean, "analytic") ~ "Statistics & Data Science",
      
      str_detect(major_clean, "computer") |
      str_detect(major_clean, "cs") |
      str_detect(major_clean, "software") ~ "Computer Science / IT",
      
      str_detect(major_clean, "engineer") ~ "Engineering (All Types)",
      
      str_detect(major_clean, "math") ~ "Mathematics",
      
      str_detect(major_clean, "bio") |
      str_detect(major_clean, "life") |
      str_detect(major_clean, "chem") |
      str_detect(major_clean, "enviro") ~ "Biological / Physical Sciences",
      
      str_detect(major_clean, "business") |
      str_detect(major_clean, "finance") |
      str_detect(major_clean, "account") |
      str_detect(major_clean, "econ") ~ "Business / Economics",
      
      str_detect(major_clean, "psych") ~ "Psychology",
      str_detect(major_clean, "edu") ~ "Education",
      str_detect(major_clean, "nurs") | str_detect(major_clean, "health") ~ "Health Sciences",
      
      TRUE ~ "Other / Undeclared"
    )
  )

# ---- 2) Summarize counts and percentages ----
major_summary <- cleaned_majors %>%
  count(major_group, name = "Count") %>%
  mutate(
    Percent = Count / sum(Count),
    Percent_Label = paste0(round(Percent * 100, 1), "%")
  ) %>%
  arrange(desc(Count))

# ---- 3) Clean, publication-ready table ----
kable(
  major_summary,
  caption = "Q4: Student Majors (Consolidated Categories)",
  col.names = c("Major Category", "Count", "Percent", "Percent Label"),
  digits = 1,
  align = "lccr"
) %>%
  kable_styling(full_width = FALSE, position = "center", font_size = 13)

# ---- 4) Optional: Horizontal bar chart ----
ggplot(major_summary, aes(y = fct_reorder(major_group, Count), x = Count)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +
  geom_text(aes(label = paste0(Count, " (", round(Percent * 100, 1), "%)")),
            hjust = -0.1, fontface = "bold", size = 3.6, color = "black") +
  labs(
    title = "Q4: Student Majors (Consolidated Categories)",
    x = "Number of Students",
    y = "Major Category"
  ) +
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 35, 15)
  )

# ---- 5) (Optional) Save the figure ----
ggsave(
  filename = "figures_all/Q4_student_majors_consolidated.png",
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

```

## Demographics table Q3-6 
```{r, warning=FALSE, message=FALSE}
## Demographics table Q3-6
# 1) Replace NA in Q5 with "Graduate" and add it as a level
FilteredDF <- FilteredDF %>%
  mutate(
   # Q5 = replace_na(Q5, "Graduate Student"),
    Q5 = factor(Q5, levels = c("Freshman", "Sophomore", "Junior", "Senior", "Graduate Student")))



# 4) Q5 summary now includes former NAs as "Graduate"
FilteredDF %>% tabyl(Q5)

# 5) Q5 bar chart (no NA filter needed; includes "Graduate")
FilteredDF %>%
  count(Q5) %>%
  ggplot(aes(x = Q5, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  ggtitle("What is your classification?") +
  labs(x = "Classification", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Q6 
```{R}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(janitor)

# ---- 1) Define ordered factor levels ----
FilteredDF$Q6 <- factor(
  FilteredDF$Q6,
  levels = c("Never", "Rarely", "Sometimes", "Often", "Very Often")
)

# ---- 2) Compute frequencies and percentages ----
plot_df_q6 <- FilteredDF %>%
  filter(!is.na(Q6)) %>%
  count(Q6) %>%
  mutate(Percent = n / sum(n))

# ---- 3) Horizontal bar chart with publication theme ----
p_q6 <- ggplot(plot_df_q6, aes(y = fct_reorder(Q6, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percentage labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(plot_df_q6$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and labels
  labs(
    title = "Q6: Frequency of Using AI Technologies in Data Science Work",
    subtitle = "Includes tools such as ChatGPT, Copilot, and Google Colab",
    x = "Percentage of Respondents",
    y = "Frequency",
    caption = "Note: Percentages calculated from valid (non-missing) responses."
  ) +

  # Apply unified publication theme
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# ---- 4) Save as high-resolution PNG ----
ggsave(
  filename = "figures_all/Q6_ai_usage_frequency_horizontal.png",
  plot = p_q6,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q6)

``` 

## Q7 ## percnetage of all mentions in responses
```{r}
## FilteredDF %>% tabyl(Q7)
## needs wrangling cuz we allowed multiple responses
# Step 1: Split comma-separated responses into rows
long_Q7 <- FilteredDF %>%
  select(Q7) %>%
  filter(!is.na(Q7)) %>%  ## remove NAs
  mutate(Q7 = str_split(Q7, ",")) %>%
  unnest(Q7)

# Step 2: Clean up whitespace
long_Q7 <- long_Q7 %>%
  mutate(Q7 = str_trim(Q7))

# Step 3: Count frequencies
Q7_summary <- long_Q7 %>%
  count(Q7, sort = TRUE) %>%
  rename(Response = Q7, Frequency = n)  
# View summary
print(Q7_summary)

## Frequency of use case as percent of 82 responses
# Step 3: Count frequencies and compute relative frequencies by total respondents (82)
Q7_summary2 <- long_Q7 %>%
  count(Q7, sort = TRUE) %>%
  rename(Response = Q7, Frequency = n) %>%
  mutate(Relative_Freq = round(Frequency / 120*100, digits=1))

# View summary
print(Q7_summary2)


```

## Q7 Percents based on valid responses
```{r}
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(scales)
library(stringr)

# ---- 1) Split and clean comma-separated responses ----
long_Q7 <- FilteredDF %>%
  select(Q7) %>%
  filter(!is.na(Q7)) %>%
  mutate(Q7 = str_split(Q7, ",")) %>%
  unnest(Q7) %>%
  mutate(Q7 = str_trim(Q7))

# ---- 2) Count responses and compute percent based on valid respondents ----
n_valid <- nrow(FilteredDF %>% filter(!is.na(Q7)))

Q7_summary <- long_Q7 %>%
  count(Q7, sort = TRUE) %>%
  rename(Response = Q7, Count = n) %>%
  mutate(
    Percent = Count / n_valid,
    Label = paste0(Count, " (", round(Percent * 100, 1), "%)")
  )

# ---- 3) Horizontal bar chart with counts + percentages ----
p_q7 <- ggplot(Q7_summary, aes(
  y = fct_reorder(Response, Percent),
  x = Percent
)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline labels showing count and percent
  geom_text(
    aes(label = Label),
    hjust = -0.1, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(Q7_summary$Percent) + 0.1),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and labels
  labs(
    title = "Q7: Common Use Cases for AI Tools in Data Science Tasks",
    subtitle = "Multiple selections allowed; percentages based on valid respondents",
    x = "Percentage of Respondents Selecting Each Use Case",
    y = "Use Case",
    caption = paste0("n = ", n_valid, 
                     " respondents (percentages may exceed 100% because multiple choices were allowed).")
  ) +

  # Minimal publication-style theme
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    plot.caption = element_text(size = 9),
    plot.margin = margin(10, 25, 25, 15)
  )

# Display chart
p_q7

```

## Q8: Are you interested in pursuing a degree in data science in the future?
```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(janitor)

# ---- 1) Define ordered factor levels ----
FilteredDF$Q8 <- factor(
  FilteredDF$Q8,
  levels = c("Yes", "No", "Not sure")
)

# ---- 2) Compute frequencies and percentages ----
plot_df_q8 <- FilteredDF %>%
  filter(!is.na(Q8)) %>%
  count(Q8) %>%
  mutate(Percent = n / sum(n))

# ---- 3) Horizontal bar chart (publication theme) ----
p_q8 <- ggplot(plot_df_q8, aes(y = fct_reorder(Q8, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percentage labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(plot_df_q8$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and labels
  labs(
    title = "Q8: Interest in Pursuing a Degree in Data Science",
    subtitle = "Responses show future interest in formal data science education",
    x = "Percentage of Respondents",
    y = "Response",
    caption = "Note: Percentages calculated from valid (non-missing) responses."
  ) +

  # Unified publication theme
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# ---- 4) Save as high-resolution PNG ----
ggsave(
  filename = "figures_all/Q8_interest_in_ds_degree_horizontal.png",
  plot = p_q8,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q8)


```

## Q9: Are you interested in getting a job in data science in the future?

```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(janitor)

# ---- 1) Define response order ----
FilteredDF$Q9 <- factor(
  FilteredDF$Q9,
  levels = c("Yes", "No", "Not sure")
)

# ---- 2) Compute frequencies and percentages ----
plot_df_q9 <- FilteredDF %>%
  filter(!is.na(Q9)) %>%
  count(Q9) %>%
  mutate(Percent = n / sum(n))

# ---- 3) Horizontal bar chart with publication theme ----
p_q9 <- ggplot(plot_df_q9, aes(y = fct_reorder(Q9, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percentage labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(plot_df_q9$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and captions
  labs(
    title = "Q9: Interest in Future Careers in Data Science",
    subtitle = "Responses show majority interest in pursuing data science roles",
    x = "Percentage of Respondents",
    y = "Response",
    caption = "Note: Percentages based on valid (non-missing) responses."
  ) +

  # Unified publication theme
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# ---- 4) Save as PNG ----
ggsave(
  filename = "figures_all/Q9_interest_in_ds_career_horizontal.png",
  plot = p_q9,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q9)

```

## Q10: How familiar are you with the following AI technologies in data science? 

```{r}
library(ggplot2)
library(dplyr)
library(forcats)
library(scales)
library(tidyr)
library(grid)



# Shared Levels and Palette
levels_q10 <- c("Not at All Familiar", "Not Very Familiar", "Neutral", "Somewhat Familiar", "Very Familiar")

pal_q <- c(
  "Not at All Familiar" = "#B2182B",  
  "Not Very Familiar"   = "#E67E22", 
  "Neutral"             = "#F1C40F",  
  "Somewhat Familiar"   = "#48C9B0",  
  "Very Familiar"       = "#117A65"   
)
#Labels and Levels
tech_map <- c(
  Q10_1 = "ChatGPT",
  Q10_2 = "Gemini",
  Q10_3 = "Google Colab",
  Q10_4 = "AI-powered\ndata visualization tools",
  Q10_5 = "AI-driven\nanalytics platforms"
)

# Data Preparation

df_long_q10 <- FilteredDF %>%
  pivot_longer(starts_with("Q10_"), names_to = "Item", values_to = "Response") %>%
  mutate(
    Technology = recode(Item, !!!tech_map, .default = NA_character_),
    Response   = factor(Response, levels = levels_q10)
  ) %>%
  filter(!is.na(Response), !is.na(Technology))

plot_df_q10 <- df_long_q10 %>%
  count(Technology, Response, .drop = FALSE) %>%
  group_by(Technology) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  group_by(Technology) %>%
  mutate(pct_familiar = sum(pct[Response %in% c("Somewhat Familiar", "Very Familiar")], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Technology = fct_reorder(Technology, pct_familiar, .desc = TRUE))


p_q10 <- ggplot(plot_df_q10, aes(y = Technology, x = pct, fill = Response)) +
  geom_bar(stat = "identity", color = "gray30", width = 0.7) +
  geom_text(
    aes(label = ifelse(pct > 0.04, paste0(round(pct * 100), "%"), "")),
    position = position_stack(vjust = 0.5),
    color = "black", size = 3, fontface = "bold"
  ) +
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = pal_q, guide = guide_legend(reverse = TRUE)) +
  labs(
    title = "Q10: Student Familiarity with AI Technologies in Data Science",
    subtitle = "Responses grouped by self-reported familiarity level",
    x = "Percentage of Responses",
    y = "AI Technology",
    fill = "Response",
    caption = "Note: Percentages represent the proportion of responses within each technology category."
  ) +
  theme_pub()


print(p_q10)
ggsave(
  filename = "figures_all/Q10_familiarity_horizontal_fxd.png",
  plot = p_q10,
  dpi = 700,
  width = 9.5,
  height = 6.5,  # slightly taller for legend space
  units = "in"
)



```

## Q11: Which AI tools have you used in your data science tasks? (Select all that apply) - Selected Choice

```{R}
#Define the exact expected options (based on your survey config)
Q11_options <- c(
  "ChatGPT",
  "Grammarly",
  "Claude and Copilot",
  "AI-powered visualization tools",
  "Copilot",
  "Gemini", 
  "Canva or AI assisted presentation tools",
  "Predictive analytics tools (e.g.DataRobot)",
  "None",
  "Others please specify"
)

#Extract responses and clean
long_Q11 <- FilteredDF %>%
  select(Q11) %>%
  filter(!is.na(Q11))

#Create a binary indicator for each known option
for (opt in Q11_options) {
  long_Q11[[opt]] <- str_detect(long_Q11$Q11, fixed(opt))
}

#Convert wide to long and count
Q11_summary <- long_Q11 %>%
  pivot_longer(cols = all_of(Q11_options), names_to =   "Response", values_to = "Selected") %>%
  filter(Selected) %>%
  count(Response, sort = TRUE)

print(Q11_summary)

## With Percents
Q11_summary2 <- long_Q11 %>%
  count(Q11, sort = TRUE) %>%
  rename(Response = Q11, Frequency = n) %>%
  mutate(Relative_Freq = round(Frequency / 120*100, digits=1))

# summary
print(Q11_summary2)
``` 

```{r}
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(scales)
library(stringr)

# Define expected options
Q11_options <- c(
  "ChatGPT",
  "Grammarly",
  "Claude and Copilot",
  "AI-powered visualization tools",
  "Copilot",
  "Gemini", 
  "Canva or AI assisted presentation tools",
  "Predictive analytics tools (e.g.DataRobot)",
  "None",
  "Others please specify"
)

#Extract and clean responses 
long_Q11 <- FilteredDF %>%
  select(Q11) %>%
  filter(!is.na(Q11))

#Create binary indicators for each tool
for (opt in Q11_options) {
  long_Q11[[opt]] <- str_detect(long_Q11$Q11, fixed(opt))
}

# Convert wide → long and summarize
Q11_summary <- long_Q11 %>%
  pivot_longer(
    cols = all_of(Q11_options),
    names_to = "Tool",
    values_to = "Selected"
  ) %>%
  filter(Selected) %>%
  count(Tool, name = "Count") %>%
  mutate(Percent = Count / 120)

# Polished horizontal bar chart 
p_q11 <- ggplot(Q11_summary, aes(y = fct_reorder(Tool, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percentage labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis settings
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(Q11_summary$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and captions
  labs(
    title = "Q11: AI Tools Used in Data Science Tasks",
    subtitle = "Multiple selections allowed; shows percentage of respondents selecting each tool",
    x = "Percentage of Respondents",
    y = "AI Tool",
    caption = "Note: Percentages calculated from valid responses."
  ) +

  # Apply your publication theme
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# Save as PNG
ggsave(
  filename = "figures_all/Q11_ai_tools_used_horizontal.png",
  plot = p_q11,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q11)

```

## Q12_1: To what extent do you agree with the following statements regarding AI tools in data science? 

```{r}
# Map Q12_
statement_map <- c(
  Q12_1 = "AI tools enhance \naccuracy",
  Q12_2 = "AI tools automate \nprocessing",
  Q12_3 = "AI tools provide \nnew insights",
  Q12_4 = "AI tools lack \nemotional intelligence"
)

# Long format + recode 
df_long_q12 <- FilteredDF %>%
  pivot_longer(
    cols = starts_with("Q12_"),
    names_to = "Item",
    values_to = "Response"
  ) %>%
  mutate(
    Statement = dplyr::recode(Item, !!!statement_map, .default = NA_character_),
    Response = factor(
      Response,
      levels = c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree")
    )
  ) %>%
  filter(!is.na(Response), !is.na(Statement))

# Compute proportions per statement
plot_df_q12 <- df_long_q12 %>%
  count(Statement, Response, .drop = FALSE) %>%
  group_by(Statement) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

#  Order statements by % favorable (Agree + Strongly Agree) 
plot_df_q12 <- plot_df_q12 %>%
  group_by(Statement) %>%
  mutate(pct_fav = sum(pct[Response %in% c("Agree", "Strongly Agree")], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Statement = fct_reorder(Statement, pct_fav, .desc = TRUE))

#Custom palette 
pal_q12 <- c(
  "Strongly Disagree" = "#B2182B",
  "Disagree"          = "#E67E22",
  "Neutral"           = "#F1C40F",
  "Agree"             = "#48C9B0",
  "Strongly Agree"    = "#117A65"
)

# Horizontal stacked proportion bar chart 
p_q12 <- ggplot(plot_df_q12, aes(y = Statement, x = pct, fill = Response)) +
  geom_bar(stat = "identity", color = "gray30", width = 0.7) +

  # Optional percentage labels
  geom_text(
    aes(label = ifelse(pct > 0.04, paste0(round(pct * 100), "%"), "")),
    position = position_stack(vjust = 0.5),
    color = "black", size = 3, fontface = "bold"
  ) +

  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = pal_q12, guide = guide_legend(reverse = TRUE)) +

  labs(
    title = "Q12: Attitudes Toward AI Tools in Data Science",
    subtitle = "Responses grouped by agreement level",
    x = "Percentage of Responses",
    y = "Statement",
    fill = "Response",
    caption = "Note: Percentages represent the proportion of responses within each statement."
  ) +

  # saved publication theme
  theme_pub()

# High-Resolution Figure 
ggsave(
  filename = "figures_all/Q12_attitudes_horizontal_pub.pdf",
  plot = p_q12,
  dpi = 600,
  width = 9,
  height = 6,
  units = "in"
)

print(p_q12)


```

## Q13: “How often do you use AI tools for the following data science tasks? (Never, Rarely, Sometimes, Often, Always)”

```{r}

# Map Q13_* 
task_map <- c(
  Q13_1 = "Exploratory data analysis",
  Q13_2 = "Predictive modeling",
  Q13_3 = "Data visualization",
  Q13_4 = "Automating data cleaning tasks",
  Q13_5 = "Presenting findings and reports"
)

# Define levels + consistent palette 
levels_q13 <- c("Never", "Rarely", "Sometimes", "Often", "Always")
pal_q13 <- c(
  "Never"     = "#B2182B",  # red
  "Rarely"    = "#E67E22",  # orange
  "Sometimes" = "#F1C40F",  # gold
  "Often"     = "#48C9B0",  # turquoise
  "Always"    = "#117A65"   # teal green
)

#Reshape and clean
df_long_q13 <- FilteredDF %>%
  pivot_longer(
    cols = starts_with("Q13_"),
    names_to = "Item",
    values_to = "Response"
  ) %>%
  mutate(
    Task     = recode(Item, !!!task_map, .default = NA_character_),
    Response = factor(Response, levels = levels_q13)
  ) %>%
  filter(!is.na(Response), !is.na(Task))

# Calculate proportions 
plot_df_q13 <- df_long_q13 %>%
  count(Task, Response, .drop = FALSE) %>%
  group_by(Task) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

#Order tasks by % high use 
plot_df_q13 <- plot_df_q13 %>%
  group_by(Task) %>%
  mutate(pct_high = sum(pct[Response %in% c("Often", "Always")], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Task = fct_reorder(Task, pct_high, .desc = TRUE))

# Horizontal stacked bar chart
p_q13 <- ggplot(plot_df_q13, aes(y = Task, x = pct, fill = Response)) +
  geom_bar(stat = "identity", color = "gray30", width = 0.7) +

  # Optional: add centered % labels inside bars
  geom_text(
    aes(label = ifelse(pct > 0.04, paste0(round(pct * 100), "%"), "")),
    position = position_stack(vjust = 0.5),
    color = "black", size = 3, fontface = "bold"
  ) +

  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  scale_fill_manual(values = pal_q13, guide = guide_legend(reverse = TRUE)) +

  labs(
    title = "Q13: How often do you use AI tools for data science tasks?",
    subtitle = "Responses grouped by frequency of use",
    x = "Percentage of Responses",
    y = "Task",
    fill = "Response",
    caption = "Note: Percentages represent the proportion of responses within each task."
  ) +

  #publication theme 
  theme_pub()

# high-resolution figure 
ggsave(
  filename = "figures_all/Q13_utilization_horizontal_pub.pdf",
  plot = p_q13,
  dpi = 600,
  width = 9,
  height = 6,
  units = "in"
)

print(p_q13)


```

## Q14: How confident are you in explaining AI-generated outputs for data analysis tasks to peers or supervisors?

```{r}
FilteredDF$Q14 <- factor(FilteredDF$Q14,
  levels = c("Very Confident", "Confident", "Neutral", 
             "Not Very Confident", "Not at All Confident"))

# Frequency table
FilteredDF %>%
  filter(!is.na(Q14)) %>%
  tabyl(Q14)


```

```{r}
# Factor order and consistent color palette
conf_levels <- c(
  "Not at All Confident",
  "Not Very Confident",
  "Neutral",
  "Confident",
  "Very Confident"
)

conf_colors <- c(
  "Not at All Confident" = "#B2182B",  
  "Not Very Confident"   = "#E67E22",  
  "Neutral"              = "#F1C40F",  
  "Confident"            = "#48C9B0",  
  "Very Confident"       = "#117A65"   
)


plot_df_q14 <- FilteredDF %>%
  filter(!is.na(Q14)) %>%
  mutate(Response = factor(Q14, levels = conf_levels)) %>%
  count(Response, .drop = FALSE) %>%
  mutate(
    pct   = n / sum(n),
    label = scales::percent(pct, accuracy = 0.1)
  )


p_q14 <- ggplot(plot_df_q14,
                aes(y = "Confidence in \nExplaining AI Outputs", x = pct, fill = Response)) +
  geom_col(width = 0.6, color = "gray30") +

  # Percent labels inside segments (only for larger segments)
  geom_text(
    aes(label = ifelse(pct >= 0.03, label, "")),
    position = position_stack(vjust = 0.5),
    size = 3, fontface = "bold", color = "black"
  ) +

  # X-axis in percentage format
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  
  scale_fill_manual(values = conf_colors, guide = guide_legend(reverse = TRUE)) +

  # Titles, labels, caption
  labs(
    title = "Q14: Ability to Explain AI-Generated Outputs",
    subtitle = "Student self-reported confidence level",
    x = "Percentage of Responses",
    y = NULL,
    fill = "Response",
#    caption = "Note: Percentages represent the proportion of respondents selecting each confidence level."
  ) +

  theme_pub()

# high-resolution figure
ggsave(
  filename = "figures_all/Q14_explainability_horizontal_pub.pdf",
  plot = p_q14,
  dpi = 600,
  width = 8.5,
  height = 4.5,   
  units = "in"
)

print(p_q14)

```

## Q15: To what extent do you agree with the following statements regarding the future of AI in data science? (Strongly Disagree, Disagree, Neutral, Agree, Strongly Agree)

```{r}
#  map and consistent palette
statement_map_q15 <- c(
  Q15_1 = "AI tools essential \nfor future workflows",
  Q15_2 = "AI improves efficiency \nin large datasets",
  Q15_3 = "AI makes DS accessible \nto non-technical professionals",
  Q15_4 = "Overreliance on AI \nmay diminish skills"
)

levels_q15 <- c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree")
pal_q15 <- c(
  "Strongly Disagree" = "#B2182B",  
  "Disagree"          = "#E67E22",  
  "Neutral"           = "#F1C40F",  
  "Agree"             = "#48C9B0",  
  "Strongly Agree"    = "#117A65"   
)

# Reshape and clean
df_long_q15 <- FilteredDF %>%
  pivot_longer(
    cols = starts_with("Q15_"),
    names_to = "Item",
    values_to = "Response"
  ) %>%
  mutate(
    Statement = recode(Item, !!!statement_map_q15, .default = NA_character_),
    Response  = factor(Response, levels = levels_q15)
  ) %>%
  filter(!is.na(Response), !is.na(Statement))

# Calculate proportions 
plot_df_q15 <- df_long_q15 %>%
  count(Statement, Response, .drop = FALSE) %>%
  group_by(Statement) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  group_by(Statement) %>%
  mutate(pct_agree = sum(pct[Response %in% c("Agree", "Strongly Agree")], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Statement = fct_reorder(Statement, pct_agree, .desc = TRUE))


p_q15 <- ggplot(plot_df_q15, aes(y = Statement, x = pct, fill = Response)) +
  geom_bar(stat = "identity", color = "gray30", width = 0.7) +


  geom_text(
    aes(label = ifelse(pct > 0.04, paste0(round(pct * 100), "%"), "")),
    position = position_stack(vjust = 0.5),
    color = "black", size = 3, fontface = "bold"
  ) +

  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = pal_q15, guide = guide_legend(reverse = TRUE)) +

  labs(
    title = "Q15: Attitudes about the Future of AI in Data Science",
    subtitle = "Responses grouped by agreement level",
    x = "Percentage of Responses",
    y = "Statement",
    fill = "Response",
    caption = "Note: Percentages represent the proportion of responses within each statement."
  ) +

  # publication theme 
  theme_pub()

# high-resolution PNG 
ggsave(
  filename = "figures_all/Q15_attitudes_horizontal_pub.png",
  plot = p_q15,
  dpi = 600,
  width = 9,
  height = 6,
  units = "in"
)

print(p_q15)


```

## Q16: Do you think students should learn how to use AI tools for their data science careers?

```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(janitor)

# Define ordered factor levels
FilteredDF$Q16 <- factor(FilteredDF$Q16,
  levels = c(
    "Yes, it's essential",
    "Yes, but not a priority",
    "Neutral",
    "No, it's not necessary",
    "No, it may even be harmful"
  )
)

# Compute frequencies and percentages
plot_df_q16 <- FilteredDF %>%
  filter(!is.na(Q16)) %>%
  count(Q16) %>%
  mutate(Percent = n / sum(n))

# Horizontal bar chart (publication style)
p_q16 <- ggplot(plot_df_q16, aes(y = fct_reorder(Q16, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percentage labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(plot_df_q16$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and captions
  labs(
    title = "Q16: Should Students Learn AI Tools for Their Data Science Careers?",
    subtitle = "Responses indicate strong support for AI tool proficiency in DS education",
    x = "Percentage of Respondents",
    y = "Response",
    caption = "Note: Percentages calculated from valid (non-missing) responses."
  ) +

  # Unified publication theme
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# high-resolution PNG
ggsave(
  filename = "figures_all/Q16_ai_learning_importance_horizontal.png",
  plot = p_q16,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q16)

```

## Q17: To what extent do you agree with the following statement: I am willing to incorporate AI tools more into my data science projects in the future.

```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(janitor)

#Define factor order
FilteredDF$Q17 <- factor(FilteredDF$Q17,
  levels = c("Strongly Agree", "Agree", "Neutral", "Disagree", "Strongly Disagree")
)

# Compute frequency and percentages 
plot_df_q17 <- FilteredDF %>%
  filter(!is.na(Q17)) %>%
  count(Q17) %>%
  mutate(Percent = n / sum(n))

#Horizontal bar chart 
p_q17 <- ggplot(plot_df_q17, aes(y = fct_reorder(Q17, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percentage labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(plot_df_q17$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and labels
  labs(
    title = "Q17: Willingness to Incorporate AI Tools in Future Data Science Projects",
    subtitle = "Responses ordered by level of agreement",
    x = "Percentage of Respondents",
    y = "Response",
    caption = "Note: Percentages calculated from valid responses."
  ) +

  # Apply your shared publication theme
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# high-resolution PNG
ggsave(
  filename = "figures_all/Q17_integration_horizontal.png",
  plot = p_q17,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q17)

```

## Q18: To what extent are you aware of the following limitations of using AI for data science tasks?  (Very Aware, Somewhat Aware, Neutral, Not Very Aware, Not at All Aware)

```{r}
# Define recode map
limitation_map <- c(
  Q18_1 = "Biased results",
  Q18_2 = "Factually incorrect outputs",
  Q18_3 = "Lack real-world \nexperience/context",
  Q18_4 = "Need validation \nby experts",
  Q18_5 = "Data privacy risks"
)

# factor levels + consistent palette
levels_q18 <- c(
  "Not at All Aware",
  "Not Very Aware",
  "Neutral",
  "Somewhat Aware",
  "Very Aware"
)

pal_q18 <- c(
  "Not at All Aware" = "#B2182B",  
  "Not Very Aware"   = "#E67E22",  
  "Neutral"          = "#F1C40F",  
  "Somewhat Aware"   = "#48C9B0",  
  "Very Aware"       = "#117A65"   
)

# Reshape and clean
df_long_q18 <- FilteredDF %>%
  pivot_longer(
    cols = starts_with("Q18_"),
    names_to = "Item",
    values_to = "Response"
  ) %>%
  mutate(
    Limitation = recode(Item, !!!limitation_map, .default = NA_character_),
    Response   = factor(Response, levels = levels_q18)
  ) %>%
  filter(!is.na(Response), !is.na(Limitation))

#Calculate proportions
plot_df_q18 <- df_long_q18 %>%
  count(Limitation, Response, .drop = FALSE) %>%
  group_by(Limitation) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

#Order by % high awareness 
plot_df_q18 <- plot_df_q18 %>%
  group_by(Limitation) %>%
  mutate(pct_high = sum(pct[Response %in% c("Somewhat Aware","Very Aware")], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Limitation = fct_reorder(Limitation, pct_high, .desc = TRUE))

# Horizontal stacked bar chart
p_q18 <- ggplot(plot_df_q18, aes(y = Limitation, x = pct, fill = Response)) +
  geom_bar(stat = "identity", color = "gray30", width = 0.7) +

  # Optional percentage labels
  geom_text(
    aes(label = ifelse(pct > 0.04, paste0(round(pct * 100), "%"), "")),
    position = position_stack(vjust = 0.5),
    color = "black", size = 3, fontface = "bold"
  ) +

  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = pal_q18, guide = guide_legend(reverse = TRUE)) +

  labs(
    title = "Q18: Awareness of AI Limitations in Data Science",
    subtitle = "Responses grouped by awareness level",
    x = "Percentage of Responses",
    y = "AI Limitation",
    fill = "Response",
    caption = "Note: Percentages represent the proportion of responses within each limitation category."
  ) +

  theme_pub()

# high-resolution PNG 
ggsave(
  filename = "figures_all/Q18_awareness_horizontal_pub.png",
  plot = p_q18,
  dpi = 600,
  width = 9,
  height = 6,
  units = "in"
)

print(p_q18)



```


## Q19: What are your primary concerns about using AI tools in data science?
```{r}
library(stringr) 
# Create new variables (binary indicators) for each concern
FilteredDF <- FilteredDF %>% mutate( concern_accuracy = str_detect(Q19, regex("accuracy", ignore_case = TRUE)), concern_transparency = str_detect(Q19, regex("transparency", ignore_case = TRUE)), concern_privacy = str_detect(Q19, regex("privacy", ignore_case = TRUE)), concern_ethics = str_detect(Q19, regex("ethic|bias", ignore_case = TRUE)), concern_skill_loss = str_detect(Q19, regex("skill", ignore_case = TRUE)), concern_understanding = str_detect(Q19, regex("understanding", ignore_case = TRUE)), 
concern_none = str_detect(Q19, regex("don't|none", ignore_case = TRUE)) ) 
# Count totals for each concern
q19_summary <- tibble( Concern = c("Accuracy & Reliability", "Lack of Transparency", "Privacy & Security", "Ethical Concerns (Bias)", "Skill Degradation", "Difficulty Understanding Models", "Don't Know/None"),
                       Count = c( sum(FilteredDF$concern_accuracy, na.rm = TRUE), sum(FilteredDF$concern_transparency, na.rm = TRUE), sum(FilteredDF$concern_privacy, na.rm = TRUE), sum(FilteredDF$concern_ethics, na.rm = TRUE), sum(FilteredDF$concern_skill_loss, na.rm = TRUE), sum(FilteredDF$concern_understanding, na.rm = TRUE), sum(FilteredDF$concern_none, na.rm = TRUE) ) ) 
# Add percentages
q19_summary <- q19_summary %>% mutate(Percent = round(Count / 120 * 100, 1)) 

library(dplyr)
library(forcats)
library(ggplot2)
library(scales)
library(stringr)
library(tibble)

# ---- 1) Binary indicators already created ----
# (as in your code)
(n19 <- sum(!is.na(FilteredDF$Q19)))
# ---- 2) Summarize + add percentages ----
q19_summary <- tibble(
  Concern = c("Accuracy & Reliability", 
              "Lack of Transparency", 
              "Privacy & Security", 
              "Ethical Concerns (Bias)", 
              "Skill Degradation", 
              "Difficulty Understanding Models", 
              "Don't Know / None"),
  Count = c(
    sum(FilteredDF$concern_accuracy, na.rm = TRUE),
    sum(FilteredDF$concern_transparency, na.rm = TRUE),
    sum(FilteredDF$concern_privacy, na.rm = TRUE),
    sum(FilteredDF$concern_ethics, na.rm = TRUE),
    sum(FilteredDF$concern_skill_loss, na.rm = TRUE),
    sum(FilteredDF$concern_understanding, na.rm = TRUE),
    sum(FilteredDF$concern_none, na.rm = TRUE)
  )
) %>%
  mutate(Percent = Count / n19)

# Polished horizontal bar chart
p_q19 <- ggplot(q19_summary, aes(y = fct_reorder(Concern, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline % labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, size = 3.6, fontface = "bold", color = "black"
  ) +

  # Axes formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(q19_summary$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and caption
  labs(
    title = "Q19: Primary Concerns About Using AI Tools in Data Science",
    subtitle = "Top issues mentioned by respondents",
    x = "Percentage of Respondents",
    y = "Concern",
    caption = "Note: Percentages calculated from valid (non-missing) responses"
  ) +

  # Unified publication theme
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

#Save as high-resolution PNG 
ggsave(
  filename = "figures_all/Q19_primary_concerns_horizontal.png",
  plot = p_q19,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q19)
```

## Q20: Have you ever encountered issues such as hallucinations, bias, or inaccuracy while using AI tools for data science tasks?


```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(janitor)

# Define factor order
FilteredDF$Q20 <- factor(FilteredDF$Q20,
  levels = c(
    "Yes, frequently", 
    "Yes, occasionally", 
    "No, never", 
    "I never used AI tools for data science"
  )
)

# Compute frequency and percentages
n20 = sum(!is.na(FilteredDF$Q20))
plot_df_q20 <- FilteredDF %>%
  filter(!is.na(Q20)) %>%
  count(Q20) %>%
  mutate(Percent = n / n20)
  

#  Horizontal bar chart
p_q20 <- ggplot(plot_df_q20, aes(y = fct_reorder(Q20, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percent labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(plot_df_q20$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles and captions
  labs(
    title = "Q20: Encountered Issues with AI Tools in Data Science Tasks",
    subtitle = "Frequency of reported challenges when using AI tools",
    x = "Percentage of Respondents",
    y = "Response",
    caption = "Note: Percentages calculated from valid (non-missing) responses."
  ) +


  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# high-resolution PNG
ggsave(
  filename = "figures_all/Q20_ai_tool_issues_horizontal_pub.png",
  plot = p_q20,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q20)

```

## Q21: Do you think AI tools can significantly enhance the following aspects of your data science workflow?  (Strongly Disagree, Disagree, Neutral, Agree, Strongly Agree)


```{r}
#  Define recode map 
aspect_map_q21 <- c(
  Q21_1 = "Automating repetitive \ntasks",
  Q21_2 = "Enhancing predictive \nmodeling accuracy",
  Q21_3 = "Visualizing \ncomplex datasets",
  Q21_4 = "Supporting faster \ndecision-making",
  Q21_5 = "Communicating to \nnon-technical stakeholders"
)

# ---- 2) Define factor levels + consistent palette ----
levels_q21 <- c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree")
pal_q21 <- c(
  "Strongly Disagree" = "#B2182B",  
  "Disagree"          = "#E67E22",  
  "Neutral"           = "#F1C40F",  
  "Agree"             = "#48C9B0",  
  "Strongly Agree"    = "#117A65"   
)

# Reshape and clean 
df_long_q21 <- FilteredDF %>%
  pivot_longer(
    cols = starts_with("Q21_"),
    names_to = "Item",
    values_to = "Response"
  ) %>%
  mutate(
    Aspect = recode(Item, !!!aspect_map_q21, .default = NA_character_),
    Response = factor(Response, levels = levels_q21)
  ) %>%
  filter(!is.na(Response), !is.na(Aspect))

#  Calculate percentages
plot_df_q21 <- df_long_q21 %>%
  count(Aspect, Response, name = "n") %>%
  group_by(Aspect) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Compute favorable % for sorting 
plot_df_q21 <- plot_df_q21 %>%
  group_by(Aspect) %>%
  mutate(pct_fav = sum(pct[Response %in% c("Agree", "Strongly Agree")], na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Aspect = fct_reorder(Aspect, pct_fav, .desc = TRUE))

#  Horizontal stacked bar chart
p_q21 <- ggplot(plot_df_q21, aes(y = Aspect, x = pct, fill = Response)) +
  geom_bar(stat = "identity", color = "gray30", width = 0.7) +

  # Inline percentage labels
  geom_text(
    aes(label = ifelse(pct > 0.04, paste0(round(pct * 100), "%"), "")),
    position = position_stack(vjust = 0.5),
    color = "black", size = 3, fontface = "bold"
  ) +

  # Format axes
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Apply consistent color palette
  scale_fill_manual(values = pal_q21, guide = guide_legend(reverse = TRUE)) +

  # Titles and labels
  labs(
    title = "Q21: Perceived Enhancement of Data Science Workflow with AI",
    subtitle = "Responses grouped by agreement level",
    x = "Percentage of Responses",
    y = "AI-Augmented Workflow Aspect",
    fill = "Response",
    caption = "Note: Percentages represent the proportion of responses within each aspect."
  ) +

  # Apply publication theme
  theme_pub()

# high-resolution PNG 
ggsave(
  filename = "figures_all/Q21_workflow_enhancement_horizontal_pub.png",
  plot = p_q21,
  dpi = 600,
  width = 9,
  height = 6,
  units = "in"
)

print(p_q21)


```

## Q22: What concerns or challenges do you face when integrating AI tools into your data science workflow?

```{r}
library(dplyr)
library(stringr)
library(forcats)
library(ggplot2)
library(scales)

# ---- 1) Create binary indicators for each challenge ----
FilteredDF <- FilteredDF %>%
  mutate(
    q22_complexity    = str_detect(Q22, regex("complexity", ignore_case = TRUE)),
    q22_accuracy      = str_detect(Q22, regex("accuracy", ignore_case = TRUE)),
    q22_understanding = str_detect(Q22, regex("understanding", ignore_case = TRUE)),
    q22_privacy       = str_detect(Q22, regex("privacy", ignore_case = TRUE)),
    q22_jobimpact     = str_detect(Q22, regex("job", ignore_case = TRUE))
  )

# ---- 2) Summarize counts ----
q22_summary <- tibble(
  Concern = c(
    "Complexity in \nUsing AI Tools",
    "Accuracy & Reliability \nof Predictions",
    "Understanding AI Outputs",
    "Data Security & \nPrivacy Risks",
    "Impact on \nDS Job Roles"
  ),
  Count = c(
    sum(FilteredDF$q22_complexity, na.rm = TRUE),
    sum(FilteredDF$q22_accuracy, na.rm = TRUE),
    sum(FilteredDF$q22_understanding, na.rm = TRUE),
    sum(FilteredDF$q22_privacy, na.rm = TRUE),
    sum(FilteredDF$q22_jobimpact, na.rm = TRUE)
  )
)

# ---- 3) Add percentages ----
n22 = sum(!is.na(FilteredDF$Q22))  # update this with your actual N
q22_summary <- q22_summary %>%
  #mutate(Percent = Count / total_n)
mutate(Percent = Count / n22)

# ---- 4) Horizontal bar chart with refined theme ----
p_q22 <- ggplot(q22_summary, aes(y = fct_reorder(Concern, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percent labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # X-axis formatting
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(q22_summary$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles & caption
  labs(
    title = "Q22: Challenges in Integrating AI Tools into Data Science Workflows",
    subtitle = "Key challenges identified from open-ended responses",
    x = "Percentage of Students",
    y = "Challenge",
    caption = "Note: Multiple responses were allowed."
  ) +

  # Apply your publication theme
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# ---- 5) Save as high-resolution PNG ----
ggsave(
  filename = "figures_all/Q22_challenges_workflow_pub.png",
  plot = p_q22,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q22)

```

## Q23: What concerns do you have about integrating AI tools into education in general?

```{r}
# binary indicators 
FilteredDF <- FilteredDF %>%
  mutate(
    q23_undermines_value = str_detect(Q23, regex("undermine|value", ignore_case = TRUE)),
    q23_limits_social    = str_detect(Q23, regex("interact|social", ignore_case = TRUE)),
    q23_hinders_skills   = str_detect(Q23, regex("skills|teamwork|problem", ignore_case = TRUE)),
    q23_overreliance     = str_detect(Q23, regex("over[- ]?reliant", ignore_case = TRUE))
  )

# Summarize counts 
q23_summary <- tibble(
  Concern = c(
    "Undermines Value of \nUniversity Education",
    "Limits Social \nInteraction",
    "Hinders Development \nof Transferable Skills",
    "Over-Reliance on \nGenerative AI"
  ),
  Count = c(
    sum(FilteredDF$q23_undermines_value, na.rm = TRUE),
    sum(FilteredDF$q23_limits_social, na.rm = TRUE),
    sum(FilteredDF$q23_hinders_skills, na.rm = TRUE),
    sum(FilteredDF$q23_overreliance, na.rm = TRUE)
  )
)

#  percentages
total_n <- 120  
q23_summary <- q23_summary %>%
  mutate(Percent = Count / total_n)

# ---- 4) Horizontal bar chart with refined styling ----
p_q23 <- ggplot(q23_summary, aes(y = fct_reorder(Concern, Percent), x = Percent)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +

  # Inline percentage labels
  geom_text(
    aes(label = paste0(round(Percent * 100, 1), "%")),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +

  # Format axes
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(q23_summary$Percent) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +

  # Titles, labels, caption
  labs(
    title = "Q23: Concerns About Integrating AI into Education",
    subtitle = "Top concerns identified from open-ended responses",
    x = "Percentage of Students",
    y = "Concern",
    caption = "Note: Multiple responses were allowed."
    
  ) +

  # Apply publication theme 
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

#  high-resolution PNG
ggsave(
  filename = "figures_all/Q23_general_concerns_horizontal_pub.png",
  plot = p_q23,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q23)


```

## Q24: Do you believe that AI tools should complement traditional methods in data science, or replace them?
```{r}
#  Wrap labels
FilteredDF$Q24 <- str_wrap(FilteredDF$Q24, width = 35)

# Tabulate and compute percentages
plot_df_q24 <- FilteredDF %>%
  filter(!is.na(Q24)) %>%
  count(Q24, name = "n") %>%
  mutate(pct = n / total_n)

# Horizontal bar chart 
p_q24 <- ggplot(plot_df_q24, aes(y = fct_reorder(Q24, pct), x = pct)) +
  geom_bar(stat = "identity", fill = "#48C9B0", color = "gray30", width = 0.7) +
  geom_text(
    aes(label = percent(pct, accuracy = 0.1)),
    hjust = -0.15, fontface = "bold", size = 3.6, color = "black"
  ) +
  scale_x_continuous(
    labels = label_percent(accuracy = 1),
    limits = c(0, max(plot_df_q24$pct) + 0.08),
    breaks = seq(0, 1, 0.2),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    title = "Q24: Should AI Tools Complement or Replace Traditional Methods?",
    subtitle = "Distribution of open-ended responses",
    x = "Percentage of Valid Responses",
    y = "Response",
    caption = "Note: Percentages represent non-missing responses."
  ) +
  theme_pub() +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10.5, lineheight = 1.1),
    plot.margin = margin(10, 20, 25, 15)
  )

# high-resolution PNG
ggsave(
  filename = "figures_all/Q24_complement_or_replace_horizontal_pub.png",
  plot = p_q24,
  dpi = 600,
  width = 8.5,
  height = 5.5,
  units = "in"
)

print(p_q24)

```